---
title: Static plots - Assignments
output: html_document
---

### References and datasets

-   Original paper: Abbas et al. (2019) *Ecological Indicators* (<https://doi.org/10.1016/j.ecolind.2019.105524>)
-   StatQuest: PCA and other videos (<https://statquest.org/>)
-   Also check the lecture slides for L02-L07

# 1. Student ID and Name

## [Student ID]{style="color:red"}: 20995548

## [Name]{style="color:red"}: HE Bingyi

# 2. Generative AI declaration

Students may use generative AI tools during assignment, for example, to check and improve code, provided that the content is original and reflects the student's own contributions. If generative AI is used, students must declare its use in the box below. An example statement is: “*During the assignment, I used [XXXX] to [XXXX]. I subsequently reviewed and edited the code and take(s) full responsibility for the final version of the code.”*

::: textbox
During the assignment, I used ChatGPT as a supporting tool to help clarify aspects of the R code and the statistical procedures used in the analysis, such as PCA and NMDS. I also consulted it to help diagnose and understand potential bugs in my code. The AI-generated suggestions were used only as references to support my learning and debugging process. I reviewed the outputs critically, revised the code where necessary, and ensured that the final implementation reflects my own understanding and work. I take full responsibility for the final submitted version of the code.
:::

# 3. Assignments

## Github: Create a Github account

-   Create your Github account.
-   Install Git on your computer as well (the latest version = 2.53.0).
-   Not mandatory, but please read the advice about usernames in the lecture.

## Github: Create your first repository

-   Repository name must be “DASC3240_Assignment1.”
-   Make the visibility “public” so I can check it.
-   The repository must have a README and a License file.
-   Include a screenshot of your shell window showing the version of Git on your computer.

## Tree community data in Hong Kong

Go to Canvas and download `Abbas_2019_TableS4_ed1.csv` and `Abbas_2019_TableS4_ed2.csv`.

-   `Abbas_2019_TableS4_ed1.csv` includes the plot information (metadata).
-   `Abbas_2019_TableS4_ed2.csv` includes dominant tree species data in the 28 forest plots.

1.  Calculate the total abundance of the dominant tree species in each forest plot, and visualize the relationship between the total abundance and forest age. You should use `AGE_MEDIAN` column in the `Abbas_2019_TableS4_ed1.csv` file as a forest age index.

2.  Visualize the dominant tree species compositions using dimension reduction techniques. For visualization, you must use different colors for different forest age groups.

    2-1. Perform PCA to visualize the tree species composition.\
    2-2. Perform NMDS to visualize the tree species composition.\

# 3. Marking criteria and important notes

[**Please check the lecture PDF "L08_Assignment-I.pdf" provided on Canvas.**]{style="color:darkred"}

# 4. Your answers

### 1. Calculate the total abundance of the dominant tree species in each forest plot, and visualize the relationship between the total abundance and forest age.

-   You should use `AGE_MEDIAN` column in the `Abbas_2019_TableS4_ed1.csv` file as a forest age index.
-   You can calculate **the sum of each column** in `Abbas_2019_TableS4_ed2.csv` to calculate the total abundance of the dominant tree species in each plot.
-   You will combine the data from two dataframes, so please check if the column names of `Abbas_2019_TableS4_ed2.csv` and the row names (indices) of `Abbas_2019_TableS4_ed1.csv` match (by using R script, not by visual inspection!).
-   To visualize the relationship, barplot is **NOT** recommended.

```{r}
### Your R code is here
### XXXXX
library(tidyverse)
library(ggplot2)

# read data (required path style)
ed1 <- read.csv("data/Abbas_2019_TableS4_ed1.csv", row.names = 1)
ed2 <- read.csv("data/Abbas_2019_TableS4_ed2.csv", row.names = 1)

# check matching (must be done by R script, not visual)
if (all(colnames(ed2) == rownames(ed1))) {
  cat("Names match between ed2 columns and ed1 rownames.\n")
} else {
  stop("Names do NOT match between ed2 columns and ed1 rownames.")
}

# total abundance per plot (sum of each column)
total_abundance <- colSums(ed2)

# combine into one df
result_df <- data.frame(
  Plot = names(total_abundance),
  TotalAbundance = as.numeric(total_abundance),
  AGE_MEDIAN = ed1[names(total_abundance), "AGE_MEDIAN"]
)

# Create the plot showing the relationship between total abundance and forest age
ggplot(result_df,
       
       # Define x and y variables and fill boxplots by forest age group
       aes(x = factor(AGE_MEDIAN),
           y = TotalAbundance,
           fill = factor(AGE_MEDIAN))) +

  # Draw boxplots summarizing abundance distribution
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +

  # Add jittered points for individual observations
  # Points are black and slightly transparent
  geom_jitter(width = 0.15,
              size = 1.5,
              alpha = 0.6,
              color = "black") +

  # Use pastel colors for boxplots
  scale_fill_brewer(palette = "Pastel1") +

  # Add title and axis labels
  labs(
    title = "Total abundance of dominant tree species vs forest age",
    x = "Median forest age (yr)",
    y = "Total tree abundance (ind.)",
    fill = "Forest age"
  ) +

  # Gray background theme
  theme_gray(base_size = 14)

```

### 2. Visualize the dominant tree species compositions using PCA and NMDS.

There are a number of points you should consider:

-   For visualization, you must use different colors for different forest age groups. Use of different marker styles for different groups is also a good idea.
-   [**Set random seeds if your analysis includes random process(es)**]{style="color:darkred"}.
-   Before applying dimension reduction functions (e.g., `prcomp()`), please make sure that [**rows represent samples (= forest plots) and columns represent tree species**]{style="color:darkred"}. You have to transpose the data frame (In `Abbas_2019_TableS4_ed2.csv`, columns represent samples and rows represent tree species).
-   Usually we should consider whether we must perform the data preprocessing before applying dimension reduction techniques (e.g., standardization). [**In this assignment, please convert your data into compositional data using the following code**]{style="color:darkred"} `d_tree_rel <- apply(d_tree, 2, function(x) x/sum(x))` (`d_tree` represents the name of your data.frame. You should change it depending on your R object name). This means that, in this assignment, we are interested in tree community compositions, not abundances.

#### 2-1. Perform PCA to visualize the tree species composition.

After PCA, please visualize the result.

```{r}
### Your R code is here
### XXXXX
# Convert PCA scores to a data frame

# Convert species abundance data to numeric
ed2_num <- as.data.frame(lapply(ed2, as.numeric))
rownames(ed2_num) <- rownames(ed2)

# Convert to relative abundance and transpose
ed2_num_rel <- apply(ed2_num, 2, function(x) x/sum(x)) %>% t()

# Run PCA
pca_ed2_num_rel <- prcomp(ed2_num_rel, center = FALSE, scale = FALSE)
summary(pca_ed2_num_rel)

# Convert PCA scores to dataframe
pca_df <- as.data.frame(pca_ed2_num_rel$x)

# Add AGE_MEDIAN
pca_df$AGE_MEDIAN <- ed1[rownames(pca_df), "AGE_MEDIAN"]

# Flip axes if needed
pca_df$PC1 <- pca_df$PC1
pca_df$PC2 <- pca_df$PC2

# Plot PCA
ggplot(pca_df, aes(x = PC1, y = PC2,
                   color = factor(AGE_MEDIAN),
                   shape = factor(AGE_MEDIAN))) +
  geom_point(size = 1.2) +
  labs(
    title = "PCA",
    x = "PC1 (49.7%)",
    y = "PC2 (18.8%)",
    color = "AGE_MEDIAN",
    shape = "AGE_MEDIAN"
  ) +
  theme_gray(base_size = 14)
```

#### 2-2. Perform NMDS to visualize the tree species composition.

After NMDS, please visualize the result. You can use Bray-Curtis dissimilarity to calculate the dissimilarity among plots.

```{r}
### Your R code is here
### XXXXX
library(vegan)
library(ggplot2)

# convert species abundance to numeric
ed2_num <- as.data.frame(lapply(ed2, as.numeric))
rownames(ed2_num) <- rownames(ed2)

# convert to relative abundance
d_tree_rel <- apply(ed2_num, 2, function(x) x/sum(x))

set.seed(123)

# run NMDS
nmds_res <- metaMDS(t(d_tree_rel), distance = "bray", k = 2, trymax = 100)

# create dataframe
nmds_df <- data.frame(
  NMDS1 = nmds_res$points[,1],
  NMDS2 = nmds_res$points[,2],
  Plot = rownames(nmds_res$points),
  AGE_MEDIAN = ed1[rownames(nmds_res$points), "AGE_MEDIAN"]
)

# plot
ggplot(nmds_df, aes(x = NMDS1, y = NMDS2,
                    color = factor(AGE_MEDIAN),
                    shape = factor(AGE_MEDIAN))) +
  geom_point(size = 1.2) +
  labs(
    title = paste0(
      "NMDS of dominant tree species composition (Bray-Curtis), stress = ",
      round(nmds_res$stress, 3)
    ),
    x = "Axis 1",
    y = "Axis 2",
    color = "AGE_MEDIAN",
    shape = "AGE_MEDIAN"
  ) +
  theme_minimal()
```
